---
title: "Exploratory exposure-response analysis"
subtitle: "The importance of understanding the scientific context"
author: 'Author: Marc Vandemeulebroecke et al'
date:  "`r format(Sys.Date(), '%A, %B %d, %Y') `"
output: html_document
---

Authors:
Marc Vandemeulebroecke, Mark Baillie, Alison Margolskee, Baldur Magnusson

```{r initialisation, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)

###################################
## Libraries 
###################################
library('knitr')
library(tidyverse)
library(RxODE)
library(caTools)
library(rstanarm)

theme_set(theme_gray()) # undo rstan theme setting

library(ggplot2)
library(dplyr)
library(gridExtra)
library(data.table)
library(grid)
library(RColorBrewer)
library(lme4)
library(emmeans)


###################################
## Path to figure directory
###################################
fig_path = 'figures/'

##################################
## Set default theme             
##################################
source('utils/paper_theme.R')

##################################
## Image size file defaults
##################################
d_width <- 116.9
d_height <- 82.7
d_dpi <- 1000

##################################
## Set seed 
##################################
set.seed(12)
```

## Purpose
To study the effect of an inhaled drug on lung function.

## Exploratory Analysis
Initial plot lung function (response variable) against plasma concentration of drug (exposure in terms of Area Under Curve).

```{r load-data}
## Read in data
my_data <- 
  read_csv("./data/401_case1_PKPDdataset_ard.csv") %>%
  filter(CYCLE == 1)
# glimpse(my_data)
```


```{r common-practice, warning=FALSE}

## Plot response vs exposure
my_data %>%
  ggplot(aes(x = AUC, y = sCHG)) + 
  geom_point() + 
  scale_y_continuous(breaks = seq(-800, 800, 200)) +  
  theme_gray(base_size = 10) +
  labs(x = "RESN", y = "LIDV", title = "", caption = "Figure 1")
```

## Problems with this plot
+ Exploratory plots often give variable names directly from source data.   
Better to give more informative description.  
For this example RESN = AUC0-24h (h*ug/mL) and  
LIDV = FEV1 change from baseline (mL).  
+ Most points are clustered at low end of x-axis
+ Need better display of gridlines
+ Annotation
+ font size

Figure 2 addresses these issues.  It also includes a LOESS smoother [^1], we see a positive non-linear trend, suggesting a shallow sigmoidal exposure-response relationship.

```{r figure2, warning=FALSE}
lbr <- scales::trans_breaks("log10", function(x) 10^x)
llb <- scales::trans_format("log10", scales::math_format(10^.x))

my_data %>%
  ggplot(aes(x = AUC, y = sCHG)) + 
  geom_point(alpha = 0.7) + 
  geom_smooth(method = "loess", colour = "red") +
  scale_x_log10(breaks = lbr, labels = llb) +     # more appropriate scale
  scale_y_continuous(breaks = seq(-800, 800, 200)) +
  annotation_logticks(sides = "b") +              # clearer scale
  labs(    # informative labels
    x = expression(paste("AUC0-24h (h*",mu,"g/mL)", sep = "")),
    y = "FEV1 change from baseline (mL)", 
        # Title is either a quesion or an answer
    title = "Exposure is positively associated with response",
    subtitle = "Loess smoother (95% CI)", 
    caption = "Figure 2"
    ) +
  paper_theme()    # clearer grid
```

At this point, an analyst may focus upon making the graph more appealing.   
This is an example of [selective attention](https://www.youtube.com/watch?v=vJG698U2Mvo), focusing on the detail but overlooking the higher purpose of the task (i.e. the “why”). Instead, let us now take a step back and revisit this example in the context of the first law of visual communication: have a clear purpose.

## Scientific question of interest

The scientific question of interest is to study the effect of a drug on 
lung function.  The implicit assumption here is causality.  However, plasma 
concentration is unlikely to instigate change in lung function.  

Better way to address the scientific question of interest.

Consider the effect of dose amount on both plasma concentration and lung function.
The expectation is for a positive correlation between dose amount and plasma 
concentration.  In addition, there would be a range of levels of plasma concentrations for a specific dose.  But what about the effect on lung function?

```{r}
my_data %>%   # take account of dose by distinguishing the observations
  ggplot(aes(x = AUC, y = sCHG, colour = factor(DOSE))) + 
  geom_point(alpha = 0.5) + 
  geom_smooth(method = "lm", se = FALSE) + 
  scale_colour_brewer(palette = "Set2" , name = "Dose (mg)") + 
  scale_x_log10(breaks = lbr, labels = llb) + 
  scale_y_continuous(breaks = seq(-800, 800, 200)) + 
  annotation_logticks(sides = "b") +
  labs(
    x = expression(paste("AUC0-24h (h*", mu, "g/mL)", sep = "")), 
    y = "FEV1 change from baseline (mL)", 
    title = "Exposure is not a better predictor of response than dose",
    caption = "Figure 3") + 
  paper_theme() + 
  theme(
    legend.position = c("right"),
    legend.title = element_text(size = 10)
    )
```

Visualization of exposure and response within levels of dose. The scatterplot is fundamentally changed by revisiting the question of interest and then applying good graphical principles.

This shows the unexpected result that response decreases for increase in exposure (plasma conc) for a given dose level

In addition, the reason that there seemed to be a positive correlation or non-linear trend is that dose amount influenced both plasma concentration and lung function.  Dose amount is a confounder variable.



[^1] Robust Locally Weighted Regression and Smoothing Scatterplots. 
Cleveland, W.S., 1979. Journal of the American Statistical Association, Vol 74(368), pp. 829-836. Taylor & Francis. DOI: 10.1080/01621459.1979.10481038


